The Decreasing Orbital Period of WASP-4\lowercase{b}

### 2018-12-06 paper/ms.tex
minor update
improved text; added companion dynamics limits

\author[0000-0002-0514-5538]{L. G. Bouma}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
\author[0000-0002-4265-047X]{J. N. Winn}
\affiliation{
    Department of Astrophysical Sciences,
    Princeton University,
    4 Ivy Lane, Princeton, NJ 08540, USA}
%-------------------------------------
% TESS Mission Architects:
% These authors should be listed in this order
% see https://spacebook.mit.edu/pages/viewpage.action?pageId=24543276
%-------------------------------------
%\\ \textcolor{red}{6 TESS Mission Architects:}
\author{G. R. Ricker} % grr@space.mit.edu WAITING
\affiliation{
  Department of Physics and Kavli Institute for Astrophysics and Space
  Research, Massachusetts Institute of Technology, Cambridge, MA
  02139, USA }
\author[0000-0001-6763-6562]{R. Vanderspek} % roland@space.mit.edu WAITING
\affiliation{
  Department of Physics and Kavli Institute for Astrophysics and Space
  Research, Massachusetts Institute of Technology, Cambridge, MA
  02139, USA }
\author[0000-0001-9911-7388]{D. W.~Latham} % dlatham@cfa.harvard.edu WAITING
\affiliation{
  Harvard-Smithsonian Center for Astrophysics, 60 Garden Street,
  Cambridge, MA 02138, USA }
\author{S. Seager} % seager@mit.edu WAITING
\affiliation{
  Department of Earth, Atmospheric, and Planetary Sciences,
  Massachusetts Institute of Technology, Cambridge, MA 02139, USA }
\author[0000-0002-4715-9460]{J. M.~Jenkins} % jon.jenkins@nasa.gov WAITING
\affiliation{
  NASA Ames Research Center, Moffett Field, CA 94035, USA }
%-------------------------------------
% 3 representatives of each of SPOC, POC, TSO, for a total of 9. 
%These 9 authors should be listed in alphabetical order
%-------------------------------------
%\\ \textcolor{red}{3 TSO coauthors:}
\author{TSO1}
\author{TSO2}
\author{TSO3}
%\\ \textcolor{red}{3 SPOC coauthors:}
\author{SPOC1}
\author{SPOC2}
\author{SPOC3}
%\\ \textcolor{red}{3 POC coauthors:}
\author{POC1}
\author{POC2}
\author{POC3}

\begin{abstract}
  %TODO: final numbers
  Combining data from the Transiting Exoplanet Survey Satellite (TESS)
  with previous studies, we show that the transit times of WASP-4b are
  incompatible with a constant orbital period. 
  In particular, the transits seem to have arrived $77.8 \pm 10.7$
  seconds early, and the period appears to be shrinking by
  $\dot{P}=-12.1 \pm 1.2$ milliseconds per year.   
  From TESS observations of WASP-6b, WASP-18b, and WASP-46b, we show
  that a systematic offset between the TESS time stamps and the barycentric
  reference sufficient to explain WASP-4b is ruled out at 6.3$\sigma$.
  If the timing variations are astrophysical,
  major contributors to the period change could include either
  apsidal precession, or tidal decay.
  The Doppler shift from WASP-4's acceleration towards us could
  account for at most one quarter of the observed period change
  (at $2\sigma$).
  Further transit and occultation studies will help confirm the
  reality of the timing variation, and eventually determine its cause.
\end{abstract}

@@

Long-term monitoring of hot Jupiter transit and occultation times
should eventually reveal two distinct processes.  First, whenever the
summed spin and orbital angular momenta in a hot Jupiter system are
below a particular critical value, the system will be unstable to
tidal decay \citep{counselman_outcomes_1973,hut_stability_1980}.  Most
hot Jupiters satisfy this condition, so their orbits should be
shrinking \citep{levrard_falling_2009,matsumura_tidal_2010}.  The rate
of decay is uncertain, and depends on how friction dissipates energy
carried by gravitational tides (\citealt{ogilvie_tidal_2014} gives a
review).  If we could directly measure orbital decay rates, it would
inform our understanding of tidal dissipation.  This is important
because the dissipation rates determine the fate of the system.
During its inspiral, if the planet fills its Roche lobe below the
stellar surface, the drag force from the stellar envelope accelerates
the inspiral, leading to a `direct-impact' merger
\citep{metzger_optical_2012,macleod_planetary_2018} Alternatively, if
the planet fills its Roche lobe above above the stellar surface, then
it could either be dynamically disrupted into an accretion disk or it
could slowly transfer mass to the star \citep{metzger_optical_2012}.
More broadly, the outcome of the eventual interaction between the
Earth and red giant Sun also hinges on the efficiency of tidal
dissipation \citep{rasio_tidal_1996}.

The second process that long-term timing studies should reveal is
rotation of the orbital ellipse within the orbital plane (apsidal
precession).  This effect has been studied extensively in eclipsing
binaries \citep[{\it e.g.},][]{russell_notes_1939,
schwarzschild_structure_1958}, and has been explored theoretically for
transiting planets
\citep{heyl_using_2007,pal_periastron_2008,jordan_observability_2008,ragozzine_probing_2009}.
However, apsidal precession has yet to be directly observed in hot
Jupiters, and the orbits would need non-zero eccentricity for it to be
detectable.  Though the eccentricities of hot Jupiters are expected to
be small, there are many mechanisms that could pump them away from
circularity (see \S~\ref{sec:apsidal_precession}, and the
possibilities considered by \citealt{bailey_understanding_2019}).  For
hot Jupiters, the apsidal precession is usually dominated by the
quadrupolar distortion of the planet by the star's tidal force
\citep{ragozzine_probing_2009}.  If we measured apsidal precession in
a hot Jupiter, we thus could convert the observed precession rate into
the planet's Love number, giving us a third parameter along with the
mass and radius with which to describe the interior structure of the
planet \citep[{\it e.g.},][who performed a similar procedure for
HAT-P-13b]{batygin_determination_2009}.  Measuring small
eccentricities and understanding their origin would also help us
understand the dynamical histories and interior structures of hot
Jupiters.  \citep[{\it
e.g.},][respectively]{dawson_origins_2018,ibgui_tidal_2010}

% Stars hosting hot Jupiters tend to spin faster than other stars of
% similar age and mass
% \citep{penev_hats-18_2016,penev_empirical_2018}.  This is expected
% because when the orbital frequency is larger than the spin rotation
% frequency, tides donate angular momentum from the planet to the
% star.  The reverse process occurs between the Earth and Moon,
% leading to the Moon's slowly increasing separation from Earth of
% $X.XX$cm~yr$^{-1}$ (CITE lunar ranging experiment).  The hot Jupiter
% star-planet separation distribution is best reproduced by models
% that account for tides~\citep{cameron_hierarchical_2018}.

The most convincing signs seen to date of either orbital decay or apsidal
precession have been in WASP-12b.
\citet{maciejewski_departure_2016} showed that the transit times of
WASP-12b did not follow a linear ephemeris at $5\sigma$ confidence,
and measured $\dot{P} = -26 \pm 4\,{\rm ms}\,{\rm yr}^{-1}$.
\citet{patra_2017} presented new transit and occultation times for
the system, agreed that curvature was present, and found
$\dot{P} = -29 \pm 3\,{\rm ms}\,{\rm yr}^{-1}$.
Assuming that the period change was caused entirely by tidal decay,
\citet{patra_2017} found values for the stellar tidal quality factor
of $Q_\star' \approx 2\times10^5$.  This level of dissipation cannot be
explained by standard equilibrium tide
models~\citep{penev_tidal_2011,ogilvie_tidal_2014}.
\citet{weinberg_tidal_2017} found though that if WASP-12 were a
subgiant, tidal dissipation due to a dynamical tide might be
sufficient to produce the required level of friction.
\cite{bailey_understanding_2019} concurred, but noted that the
observational constraints on WASP-12 itself favor a main-sequence
star, not a subgiant.  Thus, more data are needed to determine the
cause of the timing variations in WASP-12b.

% Tidal decay of hot Jupiters has been invoked to explain certain
% properties of the ensemble of star-planet systems. For example, the
% scarcity of gas giants with periods less than a day is suggestive of
% orbital decay \citep[see,
%   e.g.][]{Jackson+2008,Hansen2010,Penev+2012,Ogilvie2014}. The
% anomalously rapid rotation of some hot-Jupiter host stars has been
% attributed to transfer of the planet's orbital angular momentum
% \citep{Penev+2016}. The absence of hot Jupiters around subgiant stars
% may be caused by an acceleration of orbital decay when a star leaves
% the main sequence
% \citep{VillaverLivio2009,Hansen2010,SchlaufmanWinn2013}.  Tidal decay
% might also be responsible for the lower occurrence of close-in planets
% around rapidly rotating stars \citep{TeitlerKonigl2014}, or the
% realignment of stars and their planetary orbits
% \citep{MatsakosKonigl2015}. However, direct evidence for orbital decay
% has been lacking: there have been no clear demonstrations of a
% long-term period decrease due to orbital decay \citep[see,
%   e.g.,][]{Hoyer+2016,Wilkins+2017}.
%

This work focuses on WASP-4b, which is also a plausible candidate for
orbital decay.  The planet has radius $1.37 \pm 0.04\ R_{\rm Jup}$,
and mass $1.29 \pm 0.10\ M_{\rm Jup}$
\citep{southworth_homogeneous_2011}.  It has an orbital period of 1.34
days, orbits its host star at a distance of $a/R_\star = 5.41$, and
has been frequently observed over the past decade
\citep{wilson_wasp-4b_2008,huitson_gemini_2017}.  The orbit is
consistent with circular \citep[][but see
\S~\ref{sec:apsidal_precession}]{beerer_secondary_2011,husnoo_observational_2012,bonomo_gaps_2017}.
The host star is a G7V dwarf, with $R_\star = 0.91 \pm 0.12\ R_\odot$,
$M_\star = 0.94 \pm 0.09\ M_\odot$, $T_{\rm eff} = 5436 \pm 34\ {\rm
K}$, and an isochronal age of roughly $7\ {\rm Gyr}$
\citep{southworth_homogeneous_2011,petrucci_no_2013}.

@@

\subsection{Observations and cleaning}
WASP-4 was observed in TESS camera 2 from August 23, 2018 to September
20, during the spacecraft's second sector of science operations.  The
star is designated by the TESS Input
Catalog~\citep[TIC;][]{stassun_TIC_2018} as TIC 402026209.

The TESS spacecraft recorded images of WASP-4 in an $11\times11$ pixel
array at a cadence of 2 minutes.  It was included on the
``short-cadence'' target list thanks to the Guest Investigator
programs of J.\ Southworth and S.\ Kane (G011112 and G011183
respectively).  After being downlinked through the Deep Space Network,
the data were processed through the Science Processing Operations
Center (SPOC) pipeline at NASA Ames~\citep{jenkins_tess_2016}.  One
important step in this processing was converting from spacecraft time
to barycentric time.  SPOC then reports timestamps in the ``BTJD''
time system, where ${\rm BTJD} = {\rm BJD} - 2457000$, and all times
are in the TDB reference (see \citealt{urban_explanatory_2012} for
explanations of time systems).  The lightcurves were then vetted and
released by the MIT TESS Science Office to the Mikulski Archive for
Space Telescopes~\citep{ricker_tess_alerts_2018}.

We begin our analysis with the Presearch Data Conditioning (PDC) light
curves from MAST.  The processing steps used to find an optimal
aperture (for WASP-4, a $3\times 3$ pixel array) and the PDC-MAP
algorithm are described by \citet{smith_kepler_apertures_2017} and
\citet{smith_kepler_PDC_2017} respectively.

Our processing steps were then as follows.  First, we removed all
points with non-zero quality flags.  This removes data that is
contaminated by coarse spacecraft pointing, cosmic rays, and other
annoyances \citep{tess_data_product_description_2018}.  We then
filtered out known bad observing windows.  The data closest to
spacecraft perigee typically show ramp-like systematics, so we clipped
out the first and last hour of both orbits.  Next, we focused on the
momentum dumps. As described in the data release
notes\footnote{\url{archive.stsci.edu/hlsps/tess-data-alerts/hlsp_tess-data-alerts_tess_phot_s01_tess_v1_release-notes.pdf}},
during the first two TESS sectors, every 2.5~days the spacecraft's
reaction wheels were reset to maintain pointing stability.  These
events were assigned the quality flags corresponding to ``Reaction
Wheel Desaturation Event'' and ``Manual Exclude''.  In the WASP-4
lightcurve, these flags were simultaneously set for 54 distinct
cadences, and there were 10 momentum dumps, averaging about 10 minutes
of flagged data per dump.  Simply out of caution, we clipped out an
additional 10 minutes before and after every momentum dump.  Before
applying any filters, we had 19737 data points. After applying all the
filters, we were left with 18165 measurements, or 92\% of the original
data.

After ``cleaning'' the data, we normalized the flux measurements by
dividing out the median flux.  We then converted the timestamps from
BTJD to BJD by adding the appropriate offset of 2457000
days~\citep{tess_data_product_description_2018}.  Many of these and
subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018} We opted not to
``flatten'' the lightcurves at this stage, as is often done with {\it
e.g.}, a median filter, spline, or gaussian process regression.  Such
procedures are asymmetric about the transit minimum and can skew the
transit time measurement, and their effect on the measurement
uncertainties is difficult to capture.

@@

\begin{figure}[t]
    \begin{center}
        \leavevmode
        \includegraphics[width=0.48\textwidth]{f2.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{
        {\bf TESS saw WASP-4b transit earlier than expected.}
        Points are differences between observed transit times and the
        constant-period ephemeris predicted by previous transit
        measurements.  Black points with circle markers are the more
        precise half of literatures times; gray points are the less
        precise half.  The blue curves represent the $\pm 1\sigma$
        credible interval around the constant-period ephemeris.  The
        binned TESS observation (red point) arrived $77.8 \pm 10.7\
        {\rm seconds}$ before it was expected.
        \label{fig:arrived_early}
    }
\end{figure}

Using the times, flux measurements, and errors determined above, we
began our measurement procedure by using the Box Least Squares (BLS)
method to estimate the transit epoch, period, and duration
\citep{kovacs_box-fitting_2002}.  From the BLS parameters, we isolated
each transit to a rough estimate of its midtime, $\pm$10 transit
durations.  In each transit window, we then simultaneously fit for a
model transit, plus a local linear trend.  Our transit model uses the
analytic formulae calculated by \citet{mandel_analytic_2002}, and
implemented by \citet{kreidberg_batman_2015}.

To measure the times, we allowed four free parameters per transit: the
time of mid-transit $t_{\rm tra}$, the planet-to-star radius ratio
$R_{\rm p}/R_\star$, and the slope and intercept of the line. Our
priors were wide uniform distributions centered on the BLS estimates
for $t_{\rm tra}$ and $R_{\rm p}/R_\star$, and on a slope of zero and
an intercept of unity for the line.

We fixed the remaining transit parameters as follows.  We set the
eccentricity to zero, and the longitude of periastron to $\pi/2$.  For
the limb darkening, we assumed a fixed quadratic law, with
coefficients interpolated from the tables computed by
\citet{claret_limb_2017}.  For the period, $a/R_\star$, and
inclination, we adopted the values from~\citet{petrucci_no_2013}.
These are in reasonable agreement with the parameters reported by
\citet{gillon_improved_2009}, \citet{southworth_high-precision_2009},
and \citet{huitson_gemini_2017}.

Having defined all the free and fixed parameters, we calculated an
initial guess for the free parameters by maximizing the likelihood.
We then sampled over the posterior using the algorithm proposed by
\citet{goodman_ensemble_2010}, and implemented by
\citet{foreman-mackey_emcee_2013}.  During this process, we used the
\texttt{corner} software to check that our posteriors were being
adequately sampled~\citep{corner_2016}.

@@

% TODO: final numbers
To check that the measured uncertainties are plausible, we computed
the reduced $\chi^2$ for a linear ephemeris fit to the measured TESS
transit times.  We found that $\chi^2 = 8.4$, with 16 degrees of
freedom.  Examining the residuals by eye showed that the error
variance was being overestimated, so we multiplied the measured TESS
errors by a fudge factor $f=0.73$, forcing a reduced $\chi^2$ of
unity.  This reduced the mean transit time uncertainty from $\sim$30
seconds per transit to $\sim$22 seconds per transit.

Figure~\ref{fig:lightcurves} shows the light curves, best-fit transit
models, and residuals.  Table~1 reports the mid-transit times and
their uncertainties, and also includes the previously reported times
that are analyzed in \S~\ref{sec:timing}.

Taking the mean and standard deviation of the measured planet to star
radius ratio, we find $R_{\rm p}/R_\star=0.1538\pm0.0013$, in
agreement with the determinations by {\it e.g.},
\citet{wilson_wasp-4b_2008}, \citet{gillon_improved_2009},
\citet{winn_transit_2009}, and \citet{southworth_high-precision_2009}.

Binning the residuals to 1-hour windows and taking the median absolute
deviation, we measure a MAD of the residual lightcurves of $594\,{\rm
ppm}\,{\rm hr}^{1/2}$.  WASP-4 has a TESS-band magnitude of $T=11.78$
\citep{stassun_TIC_2018}, corresponding to a predicted photon-counting
noise\footnote{github.com/lgbouma/tnm} of $410\,{\rm ppm}\,{\rm
hr}^{1/2}$.  At $T=11.78$, with a 9-pixel aperture read noise is
expected to contribute an additional $202\,{\rm ppm}\,{\rm hr}^{1/2}$
in quadrature.  The predicted zodiacal background contribution is
$673\,{\rm ppm}\,{\rm hr}^{1/2}$, based on the zodiacal background
model calculated by \citet{winn_photonflux_2013} and used in the
\citet{Sullivan_2015} planet detection simulations.  The predicted
level of zodiacal noise seems to have been an overestimate, given the
observed RMS of WASP-4.  We leave a broader assessment of TESS's
photometric performance to future work.

@@

The median parameters and their standard deviations are reported for
all the models in Table~3.  In summary, a linear ephemeris does not
describe the observed times.  Both orbital decay and precession can
fit the observed data.  However, the precession model has two extra
free parameters, and is statistically disfavored.  Further
observations will help discriminate between orbital decay and
precession.

@@

For lack of empirical data, we now turn to theoretical mechanisms
that could maintain small eccentricities.

>>>>

\author[0000-0002-0514-5538]{L. G. Bouma}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}

\author[0000-0002-4265-047X]{J. N. Winn}
\affiliation{ Department of Astrophysical Sciences, Princeton
University, 4 Ivy Lane, Princeton, NJ 08540, USA}

\author[0000-0002-6939-9211]{T. Daylan}
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }

\author[0000-0002-3481-9052]{K. G. Stassun}
\affiliation{Vanderbilt University, Department of Physics \& Astronomy,
6301 Stevenson Center Lane, Nashville, TN 37235, USA}
\affiliation{Fisk University, Department of Physics, 1000 17th Avenue
N., Nashville, TN 37208, USA}

\author[0000-0002-7084-0529]{S. Kane}
\affiliation{Department of Earth Sciences, University of California,
Riverside, CA 92521, USA}

%-------------------------------------
% TESS Mission Architects:
% These authors should be listed in this order
% see https://spacebook.mit.edu/pages/viewpage.action?pageId=24543276
%-------------------------------------
%\\ \textcolor{red}{6 TESS Mission Architects:}
\author{G. R. Ricker} % grr@space.mit.edu WAITING
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\author[0000-0001-6763-6562]{R. Vanderspek} % roland@space.mit.edu WAITING
\affiliation{ Department of Physics and Kavli Institute for Astrophysics
and Space Research, Massachusetts Institute of Technology, Cambridge, MA
02139, USA }
\author[0000-0001-9911-7388]{D. W.~Latham} % dlatham@cfa.harvard.edu WAITING
\affiliation{ Harvard-Smithsonian Center for Astrophysics, 60 Garden
Street, Cambridge, MA 02138, USA }
\author{S. Seager} % seager@mit.edu WAITING
\affiliation{ Department of Earth, Atmospheric, and Planetary Sciences,
Massachusetts Institute of Technology, Cambridge, MA 02139, USA }
\author[0000-0002-4715-9460]{J. M.~Jenkins} % jon.jenkins@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
%-------------------------------------
% 3 representatives of each of SPOC, POC, TSO, for a total of 9. 
%These 9 authors should be listed in alphabetical order
%-------------------------------------
%\\ \textcolor{red}{3 TSO coauthors:}
\author{TSO1}
\author{TSO2}
\author{TSO3}
%\\ \textcolor{red}{3 SPOC coauthors:}
\author{J. D. Twicken} % Joseph.Twicken@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
\author{B. Wohler} % bill.wohler@nasa.gov WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
\author{D. A. Caldwell} % douglas.Caldwell@nasa.gov % WAITING
\affiliation{ NASA Ames Research Center, Moffett Field, CA 94035, USA }
\affiliation{ SETI Institute, Mountain View, CA, 94043}
%\\ \textcolor{red}{3 POC coauthors:}
\author{POC1}
\author{POC2}
\author{POC3}

\begin{abstract}
  %TODO: final numbers
  Combining data from the Transiting Exoplanet Survey Satellite (TESS)
  with previous studies, we show that the transit times of WASP-4b are
  incompatible with a constant orbital period. 
  In particular, the transits seem to have arrived $77.8 \pm 10.7$
  seconds early, and the period appears to be shrinking by
  $\dot{P}=-12.1 \pm 1.2$ milliseconds per year.   
  From TESS observations of WASP-6b, WASP-18b, and WASP-46b, we show
  that a systematic offset between the TESS time stamps and the barycentric
  reference sufficient to explain WASP-4b is ruled out at 6.3$\sigma$.
  If the timing variations are astrophysical,
  major contributors to the period change could include both
  apsidal precession and tidal decay.
  The Doppler shift from WASP-4's acceleration towards us could
  account for at most one quarter of the observed period change
  (at $2\sigma$).
  Further transit and occultation studies will help confirm the
  reality of the timing variation, and eventually determine its cause.
\end{abstract}

@@

Long-term monitoring of hot Jupiter transit and occultation times should
eventually reveal two distinct processes.  First, whenever the summed
spin and orbital angular momenta in a hot Jupiter system are below a
particular critical value, the system will be unstable to tidal decay
\citep{counselman_outcomes_1973,hut_stability_1980}.  Most hot Jupiters
satisfy this condition; their orbits should be shrinking
\citep{levrard_falling_2009,matsumura_tidal_2010}.  The rate of decay is
uncertain, and depends on how friction dissipates energy carried by
gravitational tides (\citealt{ogilvie_tidal_2014} gives a review).  If
we could directly measure orbital decay rates, it would inform our
understanding of tidal dissipation.  This is important because the
dissipation rates determine the fate of the system.  During its
inspiral, if the planet fills its Roche lobe below the stellar surface,
the drag force from the stellar envelope accelerates the inspiral,
leading to a `direct-impact' merger
\citep{metzger_optical_2012,macleod_planetary_2018} Alternatively, if
the planet fills its Roche lobe above above the stellar surface, then it
could either be dynamically disrupted into an accretion disk or it could
slowly transfer mass to the star
\citep{metzger_optical_2012,valsecchi_tidally-driven_2015,jackson_tidal_2016}.
Further afield, the outcome of the eventual interaction between the
Earth and red giant Sun also hinges on the efficiency of tidal
dissipation \citep{rasio_tidal_1996}.

The second process that long-term timing studies should reveal is
rotation of the orbital ellipse within the orbital plane (apsidal
precession).  This effect has been studied extensively in eclipsing
binaries \citep[{\it e.g.},][]{russell_notes_1939,
schwarzschild_structure_1958,borkovits_eclipse_2015}, and has been
explored theoretically for transiting planets
\citep{heyl_using_2007,pal_periastron_2008,jordan_observability_2008,ragozzine_probing_2009}.
However, apsidal precession has yet to be directly observed in hot
Jupiters, and the orbits would need non-zero eccentricity for it to be
detectable.  Though the eccentricities of hot Jupiters are expected to
be small, there are many mechanisms that could pump them away from
circularity (see \S~\ref{sec:apsidal_precession}, and
\citealt{bailey_understanding_2019}).  For hot Jupiters, the apsidal
precession is usually dominated by the quadrupolar distortion of the
planet by the star's tidal force \citep{ragozzine_probing_2009}.  If we
measured apsidal precession in a hot Jupiter system, we could thus use
the observed precession rate to compute the planet's Love number, giving
a third parameter along with the mass and radius with which to describe
the planet's interior structure \citep[{\it e.g.},][who performed a
similar procedure for HAT-P-13b]{batygin_determination_2009}.  Measuring
small eccentricities and understanding their origin would also help us
understand the dynamical histories and interior structures of hot
Jupiters.  \citep[{\it
e.g.},][respectively]{dawson_origins_2018,ibgui_tidal_2010}

While numerous indirect studies have called attention to
population-level effects of tidal decay
\citep{jackson_observational_2009,hansen_calibration_2010,penev_constraining_2012,husnoo_observational_2012,matsakos_origin_2016,cameron_hierarchical_2018,penev_empirical_2018},
the most convincing direct evidence to date for either orbital decay or
apsidal precession has come from WASP-12b.
\citet{maciejewski_departure_2016} showed that the transit times of
WASP-12b did not follow a linear ephemeris at $5\sigma$ confidence, and
measured $\dot{P} = -26 \pm 4\,{\rm ms}\,{\rm yr}^{-1}$.
\citet{patra_2017} presented new transit and occultation times for the
system, and found $\dot{P} = -29 \pm 3\,{\rm ms}\,{\rm yr}^{-1}$.
Assuming that the period change was caused entirely by tidal decay,
\citet{patra_2017} found values for the stellar quality factor of
$Q_\star' \approx 2\times10^5$.  This level of dissipation cannot be
explained by standard equilibrium tide
models~\citep{penev_tidal_2011,ogilvie_tidal_2014}.
\citet{weinberg_tidal_2017} found though that if WASP-12 were a
subgiant, tidal dissipation due to a dynamical tide might be sufficient
to produce the required level of friction.
\cite{bailey_understanding_2019} concurred, but noted that the
observational constraints on WASP-12 itself favor a main-sequence star,
not a subgiant.  More data are needed to determine the cause of the
timing variations in WASP-12b.

This work focuses on WASP-4b, which is also a plausible candidate for
orbital decay.  The planet has radius $1.37 \pm 0.04\ R_{\rm Jup}$, and
mass $1.29 \pm 0.10\ M_{\rm Jup}$ \citep{southworth_homogeneous_2011}.
It has an orbital period of 1.34 days, orbits its host star at a
distance of $a/R_\star = 5.41$, and has been frequently observed over
the past decade \citep{wilson_wasp-4b_2008,huitson_gemini_2017}.  The
orbit is consistent with circular \citep[][but see
\S~\ref{sec:apsidal_precession}]{beerer_secondary_2011,husnoo_observational_2012,bonomo_gaps_2017}.
The host star is a G7V dwarf, with $R_\star = 0.92 \pm 0.13\ R_\odot$,
$M_\star = 0.92 \pm 0.07\ M_\odot$, $T_{\rm eff} = 5400 \pm 90\ {\rm
K}$, and an isochronal age of roughly $7\ {\rm Gyr}$
\citep{southworth_homogeneous_2011,petrucci_no_2013,doyle_accurate_2013}.

@@

\subsection{Observations and cleaning}
WASP-4 was observed in TESS camera 2 from August 23, 2018 to September
20, during the spacecraft's second sector of science operations.  The
star is designated by the TESS Input Catalog as TIC 402026209
\citep{stassun_TIC_2018}.

The TESS spacecraft recorded images of WASP-4 in an $11\times11$ pixel
array every 2 minutes.  It was included on the ``short-cadence'' target
list thanks to the Guest Investigator programs of J.\ Southworth and S.\
Kane (G011112 and G011183 respectively).  After being downlinked through
the Deep Space Network, the data were processed through the Science
Processing Operations Center (SPOC) pipeline at NASA
Ames~\citep{jenkins_tess_2016}.  One important step in this processing
was converting from spacecraft time to barycentric time.  SPOC then
reports timestamps in ``BTJD'', where ${\rm BTJD} = {\rm BJD} -
2457000$, BJD is the barycentric julian date, and all times are in the
TDB reference (see \citealt{urban_explanatory_2012} for explanations of
time systems).  The lightcurves were then vetted and released by the MIT
TESS Science Office to the Mikulski Archive for Space
Telescopes~\citep{ricker_tess_alerts_2018}.

We begin our analysis with the Presearch Data Conditioning (PDC) light
curves from MAST.  The steps used to find an optimal aperture (for
WASP-4, a $3\times 3$ pixel array) and the PDC-MAP algorithm are
described by \citet{smith_kepler_apertures_2017} and
\citet{smith_kepler_PDC_2017} respectively.

We then processed the lightcurve as follows.  First, we removed all points
with non-zero quality flags.  This removes data contaminated by
coarse spacecraft pointing, cosmic rays, and other annoyances
\citep{tess_data_product_description_2018}.  We then filtered out known
bad observing windows.  The data closest to spacecraft perigee typically
show ramp-like systematics, so we clipped out the first and last hour of
both orbits.  Next, we focused on the momentum dumps. As described in
the data release
notes\footnote{\url{archive.stsci.edu/hlsps/tess-data-alerts/hlsp_tess-data-alerts_tess_phot_s01_tess_v1_release-notes.pdf}},
during the first two TESS sectors, every 2.5~days the spacecraft's
reaction wheels were reset to maintain pointing stability.  These events
were assigned quality flags corresponding to ``Reaction Wheel
Desaturation Event'' and ``Manual Exclude''.  For WASP-4,
these flags were simultaneously set for 54 distinct cadences, and there
were 10 momentum dumps, averaging about 10 minutes of flagged data per
dump.  Simply out of caution, we clipped out an additional 10 minutes
before and after every momentum dump.  Before applying any filters, we
had 19737 data points. After applying all the filters, we were left with
18165 measurements, or 92\% of the original data.

After ``cleaning'' the data, we normalized the flux measurements by
dividing out the median flux.  We then converted the timestamps from
BTJD to BJD by adding the appropriate 2457000 day offset
\citep{tess_data_product_description_2018}.  Many of these and
subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018} We opted not to
``flatten'' the lightcurves at this stage, as is often done with {\it
e.g.}, a spline or gaussian process regression.  Such procedures are
asymmetric about the transit minimum and can skew the transit time
measurement without being captured by the measurement uncertainties.

@@

\begin{figure}[t]
    \begin{center}
        \leavevmode
        \includegraphics[width=0.48\textwidth]{f2.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{
      %TODO: final number
        {\bf TESS saw WASP-4b transit earlier than expected.}
        Points are differences between observed transit times and the
        constant-period ephemeris we fit to literature transits.
        Black points are the more precise half of literatures times;
        gray points are the less precise half.  The blue curves
        represent the $\pm 1\sigma$ credible interval around the
        constant-period ephemeris.  The binned TESS observation (red
        point) arrived $77.8 \pm 10.7\ {\rm seconds}$ before it was
        expected.
        \label{fig:arrived_early}
    }
\end{figure}

Using the times, flux measurements, and errors determined above, we ran
Box Least Squares (BLS) to estimate the transit epoch, period, and
duration \citep{kovacs_box-fitting_2002}.  From the BLS parameters, we
isolated each transit to within $\pm$10 transit durations of its
midtime.  In each transit window, we then simultaneously fit for a model
transit, plus a local linear trend.  Our transit model uses the analytic
formulae calculated by \citet{mandel_analytic_2002} and implemented by
\citet{kreidberg_batman_2015}.

We allowed four free parameters per transit: the
time of mid-transit $t_{\rm tra}$, the planet-to-star radius ratio
$R_{\rm p}/R_\star$, and the slope and intercept of the line.
% Our priors were wide uniform distributions centered on the BLS
% estimates for $t_{\rm tra}$ and $R_{\rm p}/R_\star$, and on a slope of
% zero and an intercept of unity for the line.
We fixed the remaining transit parameters as follows.  We set the
eccentricity to zero, and the longitude of periastron to $\pi/2$.  We
assumed a quadratic limb-darkening law, with coefficients interpolated
from the \citet{claret_limb_2017} tables.  For the period, $a/R_\star$,
and inclination, we adopted the values from~\citet{petrucci_no_2013}.
These are in reasonable agreement with the parameters reported by
\citet{gillon_improved_2009}, \citet{southworth_high-precision_2009},
and \citet{huitson_gemini_2017}.

Having defined the free and fixed parameters, we calculated an initial
guess for the free parameters by maximizing the likelihood.  We then
sampled over the posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013}.  During this process, we used the
\texttt{corner} software to check that our posteriors were being
well-sampled~\citep{corner_2016}.

@@

% TODO: final numbers
To check that the measured uncertainties are plausible, we computed the
reduced $\chi^2$ for a linear ephemeris fit to the measured TESS transit
times.  We found that $\chi^2 = 8.4$, with 16 degrees of freedom.
Examining the residuals by eye showed that the error variance was being
overestimated, so we multiplied the measured TESS errors by a factor
$f=0.73$, forcing a reduced $\chi^2$ of unity.  This reduced the mean
transit time uncertainty from $\sim$30 seconds per transit to $\sim$22
seconds per transit.

Figure~\ref{fig:lightcurves} shows the light curves, best-fit transit
models, and residuals.  Table~1 reports the mid-transit times and
their uncertainties, and also includes the previously reported times
analyzed in \S~\ref{sec:timing}.

Taking the mean and standard deviation of the measured planet to star
radius ratio, we find $R_{\rm p}/R_\star=0.1538\pm0.0013$, in
agreement with previous determinations
\citep{wilson_wasp-4b_2008,gillon_improved_2009,winn_transit_2009,southworth_high-precision_2009}.

Binning the residuals to 1-hour windows and taking the median absolute
deviation, we measure a MAD of the residual lightcurves of $594\,{\rm
ppm}\,{\rm hr}^{1/2}$.  WASP-4 has a TESS-band magnitude of $T=11.78$
\citep{stassun_TIC_2018}, corresponding to a predicted photon-counting
noise\footnote{\url{github.com/lgbouma/tnm}} of $410\,{\rm ppm}\,{\rm
hr}^{1/2}$.  At $T=11.78$, with a 9-pixel aperture read noise is
expected to contribute an additional $202\,{\rm ppm}\,{\rm hr}^{1/2}$ in
quadrature.  The predicted zodiacal background contribution is
$673\,{\rm ppm}\,{\rm hr}^{1/2}$, based on the
\citet{winn_photonflux_2013} model, which was used in the
\citet{Sullivan_2015} planet detection simulations.  The predicted level
of zodiacal noise seems to have been an overestimate, given the observed
RMS of WASP-4.  A more detailed assessment of TESS's photometric
performance is beyond the scope of this work.

@@

The median parameters and their standard deviations are reported for all
the models in Table~3.  To summarize, a linear ephemeris does not
describe the observed times.  Both orbital decay and precession can fit
the observed data.  However, the precession model has two extra free
parameters, and is statistically disfavored.  Further observations will
help discriminate between orbital decay and precession.

@@

Lacking empirical data, we now turn to theoretical mechanisms that could
produce small eccentricities.
